Off the beaten path: \\ Gaia reveals GD-1 stars perturbed off of the main stream track

### 2018-04-30 paper/gd1.tex
sec 3.2

\begin{abstract}\noindent % trust me
Tidally-disrupted globular clusters are transformed into thin, dynamically-cold
streams of stars that are extremely valuable tracers of the large- and
small-scale distribution of mass in the Galaxy.
Most stellar streams discovered around the Milky Way reside in the Galactic halo
and are therefore primarily sensitive to the dark matter distribution.
The longest and most prominent thin stream in the Milky Way --- the GD-1 stream --- has been used
to constrain the shape of the inner halo, and is a promising place to search for
signatures of dark matter subhalo interactions.
% Aims
Here we present a sample of highly-probable members of GD-1, selected to have
small parallaxes and retrograde proper motions using data from the \gaia\
mission, and \pans\ photometry consistent with an old and metal-poor population
at a distance of $\sim10~\kpc$.
This selection extends the stream by $20^\circ$, reveals a plausible location of
the progenitor, and shows significant density variations along the stream in
unprecedented detail.
In addition to several prominent gaps, for the first time we also find evidence
of stream members off the main track.
Such features are a natural result of a stream interaction with a massive
perturber, or an indication of complex internal kinematics within the
stream progenitor.
If caused by a subhalo interaction, we estimate that the detected underdensity
and the associated spur could have been induced by a perturbation of a
$\sim10^7\,\rm M_\odot$(?) object $\sim 0.5\,$Gyr ago.
\end{abstract}

@@

As a first selection of GD-1 candidate member stars, we select stars with proper
motions $-9 < \mu_{\phi_1, \odot} < -4.5~\masyr$ and $-1.7 < \mu_{\phi_2, \odot}
< 1~\masyr$, as visualized by the rectangle in the top right panel of
\figurename~\ref{fig:selection}.
\figurename~\ref{fig:selection} (top left) shows all stars the pass the
selection on proper motion, plotted in the GD-1 coordinate system.
From kinematic selection alone, the stream is identifiable as the over-density
of stars around $\phi_2 = 0$ between longitudes $-60^\circ \lesssim \phi_1
\lesssim 0^\circ$.

\begin{figure}[h]
\begin{center}
\includegraphics[width=\textwidth]{gd1_sample.pdf}
\end{center}
\caption{
    \todo{AB}
\label{fig:selection}
}
\end{figure}

To improve the contrast of the stream over the background, we cross-match the
sample to the \pans\ photometric catalog and use the photometry to further clean
the sample.
\figurename~\ref{fig:selection} (bottom right) shows the distribution of
proper-motion-selected candidate stream stars within $\left|\phi_2\right| <
1^\circ$ in a de-reddened \pans\ color-magnitude diagram (CMD).
The over-plotted isochrone (red line) is from the MESA Isochrones \& Stellar
Tracks (MIST; \citealt{Dotter:2016, Choi:2016, Paxton:2011}) and represents a
$12~\textrm{Gyr}$ old population with $\feh = -1.35$ at a distance of
$7.8~\kpc$.
We use this isochrone to motivate a polygonal selection in de-reddened $g-i$
color and apparent $g$-band magnitude, as visualized by the shaded region in the
bottom right panel of \figurename~\ref{fig:selection}.

\figurename~\ref{fig:selection} (bottom left) shows the final sample of GD-1
stream member candidates after selecting on both proper motion and photometry.
The GD-1 stream is clearly visible in the point density of individual stars (the
positions are not binned); this is the most pure view of the stream to date.
The increased contrast shows that the stream extends at least another $20^\circ$
to negative longitudes.
Several of the under-densities and gaps hinted at from photometric selection
alone (\citealt{Koposov:2010, Carlberg:2013}) appear as striking features in
this significantly cleaner density map of the GD-1 stream.
Finally, at least two new features are visible in this new view of the stream:
(1) ``the spur,'' stars above the main stream track (in $\phi_2$) between
$-40^\circ \lesssim \phi_1 \lesssim -30^\circ$, and (2) ``the blob,'' stars
below the main stream track between $-20^\circ \lesssim \phi_1 \lesssim
-10^\circ$.
We discuss each of these in more detail in \sectionname
s~\ref{sec:results}--\ref{sec:discussion} below.

@@

We extract high-confidence stream stars from the proper motion and CMD selected
sample by further selecting stars close to the main stream track.
We use these main stream track stars to study the global properties of the
stream, and find interesting substructures off of the main stream track.

@@

We extract properties of the stream from the region defined above as a function
of $\phi_1$ by computing stream properties in overlapping windows:
we set the window size to $4^\circ$ in $\phi_1$, and shift the window center by
$1^\circ$ in longitude for each measurement.
\figurename~\ref{fig:track-and-model} shows the median latitude, $\phi_2$,
stream surface density, $XX$, stream width, $XX$, and median proper motions,
$\mu_{\phi_1}$ and $\mu_{\phi_2}$, extracted in this way, and empirical scatters
for each estimated from each window (blue lines and shaded regions).
For the surface density estimate, we subtract an estimate of the background
surface density computed from the area outside of the bounded stream region.

Clear density variations, also apparent in the 2D positions of the sources,
manifest as sharp features in the surface density estimates along the stream;
these are not a result of variable dust extinction or completeness variations
from the \gaia\ scanning pattern.
First, the visible portions of the GD-1 stream are located at high Galactic
latitudes ($b > 20^\circ$), and we therefore do not expect significant dust
extinction or variations in extinction within the stream region.
For stream longitudes $-60^\circ < \phi_1 < 10^\circ$ and $-1^\circ < \phi_2 <
1^\circ$, the median $V$-band extinction is $\textrm{med}\left(A_V\right) =
0.04~\textrm{mag}$, and the dispersion in $A_V$ in this region is $\sigma_{A_V}
\approx 0.03~\textrm{mag}$, as computed from the Schlegel-Finkbeiner-Davis
extinction map (\citealt{Schlegel:1998}).
For longitudes $\phi_1 < -60^\circ$, as the stream approaches the Galactic disk,
dust extinction becomes more appreciable, and the stream stars are harder to
select apart from the background.
Second, \todo{what to say about the Gaia scanning law...}

The density variations likely represent real morphological changes along the
GD-1 stream, as have been suggested before from binned samples of
color-magnitude-selected GD-1 stars (\citealt{Carlberg:2013, DeBoer:2018}).
The deep under-densities we see from the kinematically-cleaned sample correspond
to previously reported gaps in the stream.
There is also a well-defined, sharp over-density in the surface density profile
close to $\phi_1 \approx -13^\circ$ with roughly symmetric under-densities on
either side:
this is suggestive of a progenitor system in the final stages of dissolution
(\todo{cite who}).
\todo{Table of locations, sizes, and density constrasts for the features?}

We compare the measured stream properties to a simple model for the phase-space
density of the stream, generated by only simulating the orbital evolution of
cluster stars once they are tidally stripped from the progenitor.
In detail, during simulation, star particles are released from the Lagrange
points of the cluster with a spread in position and velocity set by the mass of
the satellite and the location within the host galaxy, with some tunable scale
parameters.
We use the parametrization and adopted scale parameters of \citet{Fardal:2015},
who fit the release distribution parameters by comparing to full $N$-body
simulations across a range of mass and orbital parameters.
This method can accurately reproduce the mean track of the resulting stream, but
the density of stars along the stream depends strongly on the mass-loss history
and internal kinematics of the progenitor system.
Here we assume a constant mass-loss rate, and release stream particles uniformly
in time from each Lagrange point.

To compute initial conditions for the stream model, we fit an orbit to the
observed properties of the stream in a fixed model for the gravitational
potential of the Milky Way.
We use a three-component potential model to represent the Galaxy, consisting of
a Miyamoto-Nagai disk (\citealt{Miyamoto:1975}), a Hernquist bulge
(\citealt{Hernquist:1990}), and a spherical Navarro-Frenk-White dark matter halo
(\citealt{Navarro:1996}).
We set the disk scale-length parameters to closely match the disk model found in
\citet{Bovy:2015}, and other parameters are set to reproduce an approximately
flat rotation curve between $5 < R < 15~\kpc$ with a circular velocity at the
solar circle, $v_{\textrm{circ}, \odot} \approx 230~\kms$.

@@

\begin{figure}[h]
\begin{center}
\includegraphics[width=\textwidth]{track_observables.pdf}
\end{center}
\caption{%
\todo{APW}
\label{fig:track-and-model}
}
\end{figure}

@@

- Spur vs. blob
- Estimate velocity offset and significance
- No indication of these features in previous work

\begin{figure}[h]
\begin{center}
\includegraphics[width=\textwidth]{features.pdf}
\end{center}
\caption{%
\todo{AB}
\label{fig:features}
}
\end{figure}

\figurename~\ref{fig:features} (middle panel) again shows the proper-motion and CMD selected stream stars, with interesting features

@@

- spur & blob as related to gaps? -- classify possible explanations in this way
- if from encounter -- could get more info on the perturber than from the gaps alone, say from delta v

>>>>

\begin{abstract}\noindent % trust me
Tidally-disrupted globular clusters are transformed into thin, dynamically-cold
streams of stars that are extremely valuable tracers of the large- and
small-scale distribution of mass in the Galaxy.
Most stellar streams discovered around the Milky Way reside in the Galactic halo
and are therefore primarily sensitive to the dark matter distribution.
The longest and most prominent thin stream in the Milky Way --- the GD-1 stream --- has been used
to constrain the shape of the inner halo, and is a promising place to search for
signatures of dark matter subhalo interactions.
% Aims
Here we present a sample of highly-probable members of GD-1, selected to have
small parallaxes and retrograde proper motions using data from the \gaia\
mission, and \pans\ photometry consistent with an old and metal-poor population
at a distance of $\sim10~\kpc$.
This selection extends the stream by $20^\circ$, reveals a plausible location of
the progenitor, and shows significant density variations along the stream in
unprecedented detail.
In addition to confirming several prominent gaps detected in photometric maps of GD-1, for the first time we find evidence
of stream members off of the main track.
Such features are a natural result of a stream interaction with a massive
perturber, or an indication of complex internal kinematics within the
stream progenitor.
Velocity offset between these features and the stream is an order of magnitude larger than the variance in velocity fields of known globular cluster, thus favoring the encounter origin.
If this is confirmed, off-stream features will enable a robust estimate of the perturber's mass.
% If caused by a subhalo interaction, we estimate that the detected underdensity
% and the associated spur could have been induced by a perturbation of a
% $\sim10^7\,\rm M_\odot$(?) object $\sim 0.5\,$Gyr ago.
\end{abstract}

@@

As a first selection of GD-1 candidate member stars, we select stars with proper
motions $-9 < \mu_{\phi_1, \odot} < -4.5~\masyr$ and $-1.7 < \mu_{\phi_2, \odot}
< 1~\masyr$, as visualized by the orange rectangle in the top right panel of
\figurename~\ref{fig:selection}.
\figurename~\ref{fig:selection} (top left) shows all stars the pass the
selection on proper motion, plotted in the GD-1 coordinate system.
From kinematic selection alone, the stream is already identifiable as the over-density
of stars around $\phi_2 = 0$ between longitudes $-60^\circ \lesssim \phi_1
\lesssim 0^\circ$.

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{gd1_sample.pdf}
\end{center}
\caption{
    \todo{AB}
\label{fig:selection}
}
\end{figure}

To improve the contrast of the stream over the background, we cross-match the
sample to the \pans\ photometric catalog and use the photometry to further clean
the sample.
\figurename~\ref{fig:selection} (bottom right) shows the distribution of
proper-motion-selected candidate stream stars within $\left|\phi_2\right| <
1^\circ$ in a \pans\ color-magnitude diagram (CMD) de-reddened following \citet{Schlafly:2011}.
The over-plotted isochrone (red line) is from the MESA Isochrones \& Stellar
Tracks (MIST; \citealt{Dotter:2016, Choi:2016, Paxton:2011}) and represents a
$12~\textrm{Gyr}$ old population with $\feh = -1.35$ at a distance of
$7.8~\kpc$.
We use this isochrone to motivate a polygonal selection in de-reddened $g-i$
color and apparent $g$-band magnitude, as visualized by the shaded region in the
bottom right panel of \figurename~\ref{fig:selection}.

\figurename~\ref{fig:selection} (bottom left) shows the final sample of GD-1
stream member candidates after selecting on both proper motion and photometry.
The GD-1 stream is clearly visible in the point density of \emph{individual} stars (the
positions are not binned); this is the most pure view of the stream to date.
The increased contrast shows that the stream extends at least another $20^\circ$
to negative longitudes.
Furthermore, the stream reaches its highest surface density where it is the narrowest ($\phi_1\sim-13^\circ$), so this map might have uncovered the location of its elusive progenitor.
Several of the under-densities and gaps hinted at from photometric selection
alone (\citealt{Koposov:2010, Carlberg:2013}) appear as striking features in
this significantly cleaner map of the GD-1 stream.
Finally, at least two new features are visible in this new view of the stream:
(1) ``the spur,'' stars above the main stream track (in $\phi_2$) between
$-40^\circ \lesssim \phi_1 \lesssim -30^\circ$, and (2) ``the blob,'' stars
below the main stream track between $-20^\circ \lesssim \phi_1 \lesssim
-10^\circ$.
We discuss each of these in more detail in \sectionname
s~\ref{sec:results}--\ref{sec:discussion} below.

@@

We extract high-confidence stream stars from the proper motion and CMD selected
sample by further selecting stars close to the main stream track.
We use these main stream track stars to study the global properties of the
stream (\S\,\ref{sec:res_global}), and find interesting substructures off of the main stream track (\S\,\ref{sec:res_gap}).

@@

We extract properties of the stream from the region defined above as a function
of $\phi_1$ by computing stream properties in overlapping windows:
we set the window size to $4^\circ$ in $\phi_1$, and shift the window center by
$1^\circ$ in longitude for each measurement.
Rows 2--5 in \figurename~\ref{fig:track-and-model} show the background-subtracted
stream surface density, \todo{$\Sigma$?}, the median latitude, $\phi_2$, and median proper motions,
$\mu_{\phi_1}$ and $\mu_{\phi_2}$, extracted in this way, and the associated empirical scatters
(blue lines and shaded regions).
We estimate the local background surface density as an average of areas north and south from the bounded stream region.

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{track_observables.pdf}
\end{center}
\caption{%
\todo{APW}
\label{fig:track-and-model}
}
\end{figure}

Clear density variations, also apparent in the 2D positions of the sources,
manifest as sharp features in the surface density estimates along the stream.
These are not a result of variable dust extinction or completeness variations
from the \gaia\ scanning pattern.
First, the visible portions of the GD-1 stream are located at high Galactic
latitudes ($b > 20^\circ$), and we therefore do not expect significant dust
extinction or variations in extinction within the stream region.
For stream longitudes $-60^\circ < \phi_1 < 10^\circ$ and $-1^\circ < \phi_2 <
1^\circ$, the median $V$-band extinction is $\textrm{med}\left(A_V\right) =
0.04~\textrm{mag}$, and the dispersion in $A_V$ in this region is $\sigma_{A_V}
\approx 0.03~\textrm{mag}$, as computed from the SFD
extinction map (\citealt{Schlegel:1998}).
For longitudes $\phi_1 < -60^\circ$, as the stream approaches the Galactic disk,
dust extinction becomes more appreciable, and the stream stars are harder to
select apart from the background.
Second, each star in our sample has at least XX visibility periods, while the sample median and variance are XX and XX, respectively, guaranteeing a robust estimate of proper motions.
Variations in the number density of faint stars originating from the scanning pattern are XX times smaller than the density variations in the stream, and their spatial distribution does not map to the stream features.
% The relative ratios between different over- and under-densities might be affected by the \gaia\ completeness at the faint end, but they 
\todo{check above sentences +  fill in}

The density variations likely represent real morphological changes along the
GD-1 stream, as have been suggested before from binned samples of
color-magnitude-selected GD-1 stars (\citealt{Carlberg:2013, DeBoer:2018}).
The deep under-densities we see from the kinematically-cleaned sample correspond
to previously reported gaps in the stream.
There is also a well-defined, sharp over-density in the surface density profile
close to $\phi_1 \approx -13^\circ$ with roughly symmetric under-densities on
either side:
this is suggestive of a progenitor system in the final stages of dissolution
(\todo{cite who}).
% \todo{Table of locations, sizes, and density constrasts for the features?}

We compare the measured stream properties to a simple model for the phase-space
density of the stream, generated by only simulating the orbital evolution of
cluster stars once they are tidally stripped from the progenitor.
In detail, during simulation, star particles are released from the Lagrange
points of the cluster with a spread in position and velocity set by the mass of
the satellite and the location within the host galaxy, with some tunable scale
parameters.
We use the parametrization and adopted scale parameters of \citet{Fardal:2015},
who fit the release distribution parameters by comparing to full $N$-body
simulations across a range of mass and orbital parameters.
This method can accurately reproduce the mean track of the resulting stream, but
the density of stars along the stream depends strongly on the mass-loss history
and internal kinematics of the progenitor system.
Here we assume a constant mass-loss rate, and release stream particles uniformly
in time from each Lagrange point; these choices match results from more detailed numerical experiments of globular cluster disruption \citep[e.g.,][]{Kupper:2012}.

To compute initial conditions for the stream model, we fit an orbit to the
observed properties of the stream in a fixed model for the gravitational
potential of the Milky Way.
We use a three-component potential model to represent the Galaxy, consisting of
a disk (\citealt{Miyamoto:1975}), a bulge
(\citealt{Hernquist:1990}), and a spherical dark matter halo
(\citealt{Navarro:1996}).
We set the disk scale-length parameters to closely match the disk model found in
\citet{Bovy:2015}, and other parameters are set to reproduce an approximately
flat rotation curve between $5 < R < 15~\kpc$ with a circular velocity at the
solar circle, $v_{\textrm{circ}, \odot} \approx 230~\kms$.

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{orbitfit.pdf}
\end{center}
\caption{%
\todo{APW}
\label{fig:orbitfit}
}
\end{figure}

@@

\figurename~\ref{fig:orbitfit} shows \todo{APW: ...}

@@

- overdensities apparent beyond the main stream track used in previous section -- if truly associated with the stream, this would be a novel insight into disruption from the \gaia\ mission
- since some contamination still remains present, it is possible that this is a variation in the MW population, just randomly coming into our selection boxes
- we thus perform an experiment and compare cmd and proper motions of spur and blob fields to stream and background fields to see which they are associated with

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{features.pdf}
\end{center}
\caption{%
\todo{AB}
\label{fig:features}
}
\end{figure}

- central panel of figure~\ref{fig:features} shows the map of GD-1, with fields of interest in shaded rectangles
- we first analyze the spur, region, stream same phi1, phi2 below, and an off-stream field in phi2
- for each of these fields, we show a pair of plots in the top row: cmd on the left and proper motions on the right. 
- we first select on pm, and show cmd, and then select on cmd and show pm
- in both of these spaces, selection regions used to make the main catalog are shown in shaded orange.
- stream field (top left) has clear overdensity in the cmd selection box and clusters in proper motion
- on the other hand, the off field has some stars entering both selection boxes, but without any overdensities in either
- the spur field has fewer stars than the stream, but XX more than expected from the MW alone
- furthermore, the distribution in selection boxes is distributed similarly as in the stream field
- this confirms the association, but we note that in detail they differ
- median in pm offset from the stream field just below is XX, YY mas yr
- also dispersion different
- don't have enough stars to measure distance from the turn-off, so defer to future studies to measure by some other means, e.g., from photometry

- analogous paragraph for blob on the bottom

% - Spur vs. blob
% - Estimate velocity offset and significance
% - No indication of these features in previous work

% \figurename~\ref{fig:features} (middle panel) again shows the proper-motion and CMD selected stream stars, with interesting features

@@

- spur \& blob as related to gaps? -- classify possible explanations in this way
- if from encounter -- could get more info on the perturber than from the gaps alone, say from delta v
